---
title: "Final Project"
output: pdf_document
fontsize: 12pt
---

```{r}
getwd() # make sure that the file is in the working directory

# load the data as a data.frame
birth <- read.csv(file = "chds_births.csv")

# produce the pair plots for different combinations of  variables in the dataset

par(mar = c(4, 4, 0, 0) + .1) # size of margins

P1 <- pairs(~ wt + gestation + parity  + mage + mwt, 
      labels = c("birth weight", "gestation", "parity",
                 "mage", "mwt"), 
      pch = 19,
      cex = 0.05,
      data = birth)

P2 <- pairs(~ wt + gestation + parity + meth + mht + mwt,
      labels = c("weight", "gestation", "parity", "meth", "mht", "mwt"),
      pch = 19,
      cex = 0.05,
      data = birth)
```

```{r}
# check the relationship between income and education
pairs(~ income + med + fed, 
      labels = c("income", "med", "fed"),
      cex = .05,
      pch = 19,
      data = birth)
```

```{r}
# convert categorical variables to factors

# convert meth to non-numeric categorical variable
meth2 <- birth$meth

# "levels" of variable meth
methnames <- c("Caucasian","Mexican", "African-American","Asian", "Mixed", "Other")

# assign "Caucasian" level to values less than or equal to 5
# otherwise assign levels based on order in methnames vector
for(ii in 0:10) {
  if(ii<=5){
    meth2[meth2 == ii] <- "Caucasian"
  } else {
    meth2[meth2 == ii] <- methnames[ii-4]
  }
}

meth2 <- factor(meth2, levels = methnames) # order levels based on methnames
levels(meth2)
birth$meth <- meth2 # replace in dataset

############################################

# convert med to categorical variables
med2 <- birth$med

# "levels" of variable med
mednames <- c("elementary school", "middle school", "high school", "high school + trade", "high school + some college", "college grad", "trade school", "high school unclear")

med2 <- mednames[birth$med+1]

med2 <- factor(med2, levels = mednames) # order levels based on mednames
levels(med2)
birth$med <- med2 # replace in dataset

##########################################

# convert feth to categorical variables
feth2 <- birth$feth

# "levels" of variable feth
fethnames <- c("Caucasian","Mexican", "African-American","Asian", "Mixed", "Other")

# assign "Caucasian" level to values less than or equal to 5
# otherwise assign levels based on order in fethnames vector
for(ii in 0:10) {
  if(ii<=5){
    feth2[feth2 == ii] <- "Caucasian"
  } else {
    feth2[feth2 == ii] <- fethnames[ii-4]
  }
}

feth2 <- factor(feth2, levels = fethnames) # order levels based on methnames
levels(feth2)
birth$feth <- feth2 # replace in dataset

############################################

# convert fed to categorical variables
fed2 <- birth$fed

# "levels" of variable fed
fednames <- c("elementary school", "middle school", "high school", "high school + trade", "high school + some college", "college grad", "trade school", "high school unclear")

fed2 <- fednames[birth$fed+1]

fed2 <- factor(fed2, levels = fednames) # order levels based on fednames
levels(fed2)
birth$fed <- fed2 # replace in dataset

######################################################################

# marital
marital <- birth$marital
maritalnames <- c("married", "legally separated", "divorced", "widowed",
                  "never married") # new values
# encode the factor
marital2 <- rep(NA, length(marital))
for (ii in 1:5) {
  marital2[marital == ii] <- maritalnames[ii]
}
marital2 <- factor(marital2, levels = maritalnames)
birth$marital <- marital2
######################################################################
# income
income <- birth$income
incomenames <- c("under 2500", "2500-4999", "5000-7499", "7500-9999",
                  "10000-12499", "12500-14999", "15000-17499",
                 "17500-19999", "20000-22499", "over 22500") # new values
# encode the factor
income2 <- incomenames[income+1]
income2 <- factor(income2, levels = incomenames)
birth$income <- income2
############################################################
# smoke
smoke <- birth$smoke
smokenames <- c("never", "smokes now", "until pregnancy", 
                "used to, not anymore") # new values
# encode the factor
smoke2 <- smokenames[smoke+1]
smoke2 <- factor(smoke2, levels = smokenames)
birth$smoke <- smoke2
############################################################
# time
time <- birth$time
timenames <- c("never smoked", "still smokes", "during pregnancy", 
                "less than a year", "1-2 years", "2-3 years", "3-4 years",
               "5-9 years", "more than 10 years", 
               "quit but don't know when") # new values
# encode the factor
time2 <- timenames[time+1]
time2 <- factor(time2, levels = timenames)
birth$time <- time2
#############################################################
# number (we create new descriptive groupings for number of cigarettes smoked per day)
num2 <- birth$number
# new categories for variable number
numnames<- c("never smoked", "up to half a pack", "half to full pack", "more than one pack")

for(ii in 0:9) {
  if(ii==0){
    num2[num2 == ii] <- numnames[1]
  } 
  if(ii <= 2) {
    num2[num2 == ii] <- numnames[2]
  } 
  if(ii <= 4) {
    num2[num2 == ii] <- numnames[3]
  } else {
    num2[num2 == ii] <- numnames[4]
  }
}
num2 <- factor(num2, levels = numnames) # order levels based on numnames
levels(num2)
birth$number <- num2 # replace in dataset

```

```{r}
# check how many observations contain too many NA's
n <- length(birth$wt) # number of total observations
boundary <- .10 * n # boundary to decide if there are too many NA's

NA.list <- NA # initialize the list of required covariates

# identify and display all required covariates
for (ii in 1:length(birth)) {
 if(sum(is.na(birth[ii])) >= boundary)
 {NA.list <- c(NA.list, colnames(birth)[ii])}
}
NA.list <- NA.list[2:length(NA.list)]
NA.list
```

**Conclusion**: From our analysis, the father's height (fht), the father's weight (fwt) and the family yearly income (income) contain "too many" NA's. It implies that if we include any of these three factors into our models, then it is probable that our models will be biased because there are too many unavailable data and we do not know why they are unavailable; as a result, any model including any of these three factors will not be representative of the **entire population**. Therefore, based on the given dataset, it is reasonable not to consider these three factors. **TBC**

```{r}
# throw out fht, fwt and income
birth$fht <- NULL
birth$fwt <- NULL
birth$income <- NULL
```

```{r NA2}
# want to count how many observations contain at least one NA
n <- length(birth$wt) # number of rows
count = 0 # initialize count of rows with at least one NA
for (ii in 1:n) {
  if(sum(is.na(birth[ii,])) >= 1) {
    count = count + 1
  }
}

total_NA_rows <- count
total_NA_rows
```


```{r}
# Combine smoke and time into a single variable smoke:
newnames <- c("Never Smoked", "Still Smokes","Smoked but quit during pregnancy", "Smoked but quit before pregnancy") # create categories for new variables

# load the original data
newbirth <- read.csv(file = "chds_births.csv")

# extract the corresponding columns
smoke <- newbirth$smoke

n <- length(smoke) # number of observations
newsmoke <- rep(NA, n)

for (ii in 1:n) {
  if (is.na(smoke[ii])) {
    newsmoke[ii] <- "NA"
  }
  else {
    for (jj in 0:3) {
      if (smoke[ii] == jj) {
        newsmoke[ii] <- newnames[jj+1]
      }
    }
  }
}

# assign the new variable to the dataset
birth$smoke <- newsmoke
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```


